When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-11-08 paper/paper.tex
More changes from Christine

Throughout, our graphs are undirected, and do not have any loops. In some cases vertices have
labels, which are treated as integers; we write $\ell(v)$ for the label of vertex $v$. Two vertices
are \emph{adjacent} if there is an edge between them. The \emph{order} of a graph is the cardinality
of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a graph $G$. The
\emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the same vertex set
as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$ and $w$ are not
adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$ vertices, and an
edge between each distinct unordered pair of vertices with independent probability $p$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes
RAM, running Scientific Linux release 6.7. We will be working with four subgraph isomorphism
solvers: the Glasgow solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9 with the ``-O3'' option.

The Glasgow and LAD solvers are constraint programming inspired algorithms (although they use dedicated
data structures, and do not use a constraint programming toolkit). For each vertex in the pattern
graph, they create a \emph{variable} whose \emph{domain} contains one \emph{value} for each vertex
in the target graph. The algorithms perform backtracking search, with the goal of assigning to each
variable a value from its domain so as to form a subgraph isomorphism. The algorithms use
\emph{inference} to process \emph{constraints} which eliminate values from domains, ensuring that
the injectivity (an \emph{all-different} constraint) and adjacency rules are respected. If at any
point during search a \emph{domain wipeout} occurs (that is, if every value is removed from any
domain), the algorithms will backtrack immediately. The search is guided by \emph{variable-ordering
heuristics}, which select which variable (i.e.\ pattern vertex) to branch on next, and
\emph{value-ordering} heuristics which select which value (i.e.\ target vertex) to try as part of a
partial mapping. The two algorithms differ in terms of additional inference performed, their
choices of underlying data structures and algorithms, and in the details of their variable- and
value-ordering heuristics.

The approach used by VF2 and VF3 is a little different: domains of variables are not stored (so the
algorithm operates in the style of conventional backtracking), and wipeouts are not detected until
an assignment is made. Instead, an assignment is built up by repeatedly branching only on vertices
which are adjacent to a previously-assigned vertex.

We measure only the number of recursive calls (guessed assignments) made, not runtimes. We are not
aiming to compare absolute performance between solvers; rather, we are looking for
solver-independent patterns of difficulty. All experiments use a timeout of 1,000 seconds, which is
enough for the Glasgow solver to solve nearly all instances (whose orders were selected with this
timeout in mind), although we may slightly overestimate the proportion of unsatisfiable instances
for extremely sparse or dense pattern graphs. The LAD, VF2, and VF3 solvers experienced many more
failures with this timeout, so our picture of just how hard the hardest instances are with these
solvers is less detailed.

The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem.

@@

    \item Domain-based heuristics are much stronger than VF2's adjacency branching rules. Picking
        from small domains dynamically allows search to focus on the hardest part of the problem.

    \item This approach automatically combines features. For example, a pattern~/ target pair may
        have matching label features, and matching degree features, but if the only red pattern
        vertex has degree three whilst no red target vertex has degree more than two, domain
        filtering will detect this immediately. When combined with all-different propagation and
        maintained during search, this effect is even stronger.

    \item Finally, as indexing systems get more and more complex in an attempt to filter out a few more
        instances where VF2 performs poorly, the cost of index construction and maintenance is
        considerable.  Indices proposed in the literature often take many hours to build, and can
        consume much more space than the original graphs.
\end{itemize}

@@

\acks{The authors would like to thank Kitty Meeks and Craig Reilly for their comments, and Iva
Babukova for her help with the datasets. The heatmap schemes used in this paper were created with
the assistance of the ``Chroma.js Color Scale Helper'' by Gregor Aisch.}

>>>>

Throughout, our graphs are undirected, and do not have any loops. In some cases vertices have
labels, which are treated as integers; we write $\ell(v)$ for the label of vertex $v$. Two vertices
are \emph{adjacent} if there is an edge between them. The \emph{neighbourhood} of a vertex is the
set of vertices to which it is adjacent, and the \emph{degree} of a vertex is the cardinality of its
neighbourhood. The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\shortoverline{G}$, is the graph with the same vertex set as $G$, and with an edge between
distinct vertices $v$ and $w$ if and only if $v$ and $w$ are not adjacent in $G$. We write $G(n, p)$
for an Erd\H{o}s-R\'enyi random graph with $n$ vertices, and an edge between each distinct unordered
pair of vertices with independent probability $p$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes
RAM, running Scientific Linux release 6.7. We will be working with four subgraph isomorphism
solvers: the Glasgow solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9 with the ``-O3'' option.
The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem.

\subsection{Subgraph Isomorphism Solvers}

The Glasgow and LAD solvers are constraint programming inspired algorithms (although they use
dedicated data structures, and do not use a constraint programming toolkit). Constraint programming
is a field concerned with solving \emph{constraint satisfaction} problems. In such a problem, we
have a set of \emph{variables}, each of which has a \emph{domain} of \emph{values}. We also have a
set of \emph{constraints}. The objective is to assign each variable a value from its domain, whilst
respecting every constraint. This is done using a combination of \emph{inference}, which eliminates
infeasible values from domains, and backtracking search. This search is driven by
\emph{variable-ordering} heuristics, which select which unassigned variable to branch on, and
\emph{value-ordering} heuristics, which select which value to try giving that variable. If at any
point during search a \emph{domain wipeout} occurs (that is, if every value is removed from any
domain), the algorithms will backtrack immediately.

In the case of subgraph isomorphism, the basic constraint model is to have a domain for each pattern
vertex, with the domains being each target vertex. An \emph{all-different} constraint ensures
injectivity, whilst the adjacency rules can be expressed as extensional constraints by listing
allowed tuples.  With this model, the backtracking search builds up partial assignments of pattern
vertices to target vertices, until either a complete assignment is reached (and the instance is
\emph{satisfiable}), or search establishes that no such assignment can exist (and it is
\emph{unsatisfiable}).

This model can be extended with additional constraints. These allow more inference to be performed,
and so a solution may be found (or the lack of solution may be proven) with less search.  The
simplest example is that a pattern vertex of degree $d$ may only be mapped to a target vertex of
degree at least $d$, and so constraints can be posted which eliminate values corresponding to target
vertices of low degree from domains corresponding to pattern vertices of high degree. Richer
constraints involving neighbourhood degree sequences \citep{DBLP:journals/constraints/ZampelliDS10},
neighbourhood difference constraints \citep{DBLP:journals/ai/Solnon10}, and distances and paths
\citep{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16} are also
possible.

The Glasgow and LAD solvers both use extensions of this basic model, but differ in terms of the
additional constraints used (and hence the inference which is performed), their choices of
underlying data structures and algorithms, and in the details of their variable- and value-ordering
heuristics.

The approach used by VF2 and VF3 is different: domains of variables are not stored (so the algorithm
operates in the style of conventional backtracking), and wipeouts are not detected until an
assignment is made. Instead, an assignment is built up by repeatedly branching only on vertices
which are adjacent to a previously-assigned vertex.

Our experiments will primarily measure the number of recursive calls (guessed assignments) made, not
runtimes. We are not aiming to give a simple comparison of absolute performance between solvers;
rather, we are looking for solver-independent patterns of difficulty. All experiments use a timeout
of 1,000 seconds, which is enough for the Glasgow solver to solve nearly all instances (whose orders
were selected with this timeout in mind), although we may slightly overestimate the proportion of
unsatisfiable instances for extremely sparse or dense pattern graphs. The LAD, VF2, and VF3 solvers
experienced many more failures with this timeout, so our picture of just how hard the hardest
instances are with these solvers is less detailed.

@@

    \item Domain-based heuristics are widely studied and well-understood
        \citep{DBLP:conf/ecai/SmithG98}, and unlike VF2's adjacency branching rules, provide
        discrimination even on dense and low-diameter graphs. Picking from small domains dynamically
        allows search to focus on the hardest part of the problem, and can exploit conditional
        structure which is not visible at the top of search.

    \item Domain-based algorithms automatically combines features. For example, a pattern~/ target
        pair may have matching label features, and matching degree features, but if the only red
        pattern vertex has degree three whilst no red target vertex has degree more than two, domain
        filtering will detect this immediately. When combined with all-different propagation and
        maintained during search, this effect is even stronger.

    \item Finally, as indexing systems get more and more complex in an attempt to filter out a few more
        instances where VF2 performs poorly, the cost of index construction and maintenance is
        considerable.  Indices proposed in the literature often take many hours to build, and can
        consume much more space than the original graphs \citep{DBLP:journals/tkde/HongZLY15}.
\end{itemize}

@@

\acks{The authors would like to thank Kitty Meeks and Craig Reilly for their comments. We are
grateful to Iva Babukova for her help with the datasets and for sharing the results of her
experiments. The heatmap schemes used in this paper were created with the assistance of the
``Chroma.js Color Scale Helper'' by Gregor Aisch.}
